\documentclass[../main.tex]{subfiles}

\pagestyle{plain}
\title{\vspace{-0.5cm} Syllabus}
\author{MA 725, fall 2025}

\begin{document}
\maketitle

\subsection*{Instructor} Brian Williams. Office: CDS 411. email: \email{bwill22@bu.edu}.
\begin{itemize}
  \item Office hours: M 3:30pm--4:30pm, F 11:30pm-12:30pm.
  \item Website: \url{https://brianrwilliams.github.io/}
  \end{itemize}

  \subsection*{Time and location} 
  There are three lectures per week: MWF 2:30pm--3:20pm in KCB 201.

      \subsection*{Course summary}
      This course is about the foundations of differential geometry.
      Topics include:
\begin{enumerate}
  \item Riemannian metrics, affine connections, Levi--Civita connection.
  \item Vector bundles, general theory.
  \item Connections, (Riemannian) curvature(s).
  \item Exponential maps in Riemannian geometry and Lie theory.
  \item Characteristic classes of vector bundles.
  \item Chern, Euler, and Pontryagin.
  \item Principal bundles, connections, characteristic classes and Chern--Weil theory.
  \item Further topics.
\end{enumerate}

\subsection*{Textbook}

The main reference will be lecture notes together with \textit{Differential geometry: Connections, curvature, and
characteristic classes}, by Loring Tu.
Let me know if you are having trouble finding it.

\subsection*{Prerequisites}

The prerequisite for the class is the sequence MA 721-722 or equivalent.
If you are having trouble enrolling, and you think you meet the course requirements, please email \url{bwill22@bu.edu}.

\subsection*{Assessment}

Your grade will be determined by \textbf{five} homework assignments and a \textbf{take-home final} exam. 
Homework counts towards 50\% of your grade and the final is worth the remaining 50\% of your grade.

\subsection*{How to suceed in this class}
\begin{itemize}
  \item This class moves at a fast pace and attendance in class is necessary. 
  \item Turn your homework in. I \textbf{will not accept late homework} under any circumstances.
      For example, if you have only solved part of the homework assignment, it is in your best interest to turn in the
      work that you have done
      before the deadline.
    \item The same rule applies to the take-home final exam.
      The deadline for the final exam is final, and cannot be extended under any circumstances. (The final exam will be
      assigned 3 days prior to the last day of class, and will be due at midnight on the last day of class. Further
      details will appear as we get closer to the exam.) 
\end{itemize}


\begin{center}
  \textit{This syllabus is subject to minor changes. 
    If there is a change to the syllabus made after the first day
    of class there will be an
    announcement and it will be updated here.}
  \end{center}
\end{document}

